---
title: "Variance, Covariance and Correlation: Concepts through Simulation"
subtitle: "STAT 400 -- Multiple Random Variables"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 6.5, fig.height = 4, fig.align = "center",
                      message = FALSE, warning = FALSE)
set.seed(400)
# helpers that treat a list of equally likely outcomes as the whole population
pop_cov <- function(a, b) mean(a * b) - mean(a) * mean(b)
pop_cor <- function(a, b) pop_cov(a, b) / sqrt(pop_cov(a, a) * pop_cov(b, b))
```

# The big picture

| Quantity | Question it answers | Formula |
|---|---|---|
| **Variance** | How much does *one* variable bounce around its mean? | $\mathrm{Var}(X)=E[(X-\mu_X)^2]=E[X^2]-\mu_X^2$ |
| **Covariance** | When $X$ is above its mean, is $Y$ usually above (+) or below (−) its mean? | $\mathrm{Cov}(X,Y)=E[(X-\mu_X)(Y-\mu_Y)]=E[XY]-\mu_X\mu_Y$ |
| **Correlation** | The same, but with units removed, between $-1$ and $1$ | $\rho=\dfrac{\mathrm{Cov}(X,Y)}{\sigma_X\sigma_Y}$ |

Three ideas to keep coming back to:

1. Covariance and correlation measure **linear** togetherness only.
2. Independent $\Rightarrow$ covariance $=0$, but **not** the other way around.
3. Correlation is **not** causation.

In each section below we run an experiment, look at the numbers, and interpret them.

# Variance: one variable bouncing around

*Experiment:* the number of minutes your bus is late. Two routes have the **same average** delay (5 minutes) but different reliability.

```{r variance}
route_A <- rnorm(10000, mean = 5, sd = 1)   # reliable
route_B <- rnorm(10000, mean = 5, sd = 4)   # unpredictable
data.frame(route = c("A", "B"),
           mean = c(mean(route_A), mean(route_B)),
           variance = c(var(route_A), var(route_B)),
           sd = c(sd(route_A), sd(route_B)))
```

```{r variance-plot}
hist(route_B, breaks = 50, col = rgb(1, 0, 0, .4), xlim = c(-10, 20),
     main = "Bus delay (minutes)", xlab = "minutes late")
hist(route_A, breaks = 50, col = rgb(0, 0, 1, .4), add = TRUE)
legend("topright", c("Route A (sd 1)", "Route B (sd 4)"),
       fill = c(rgb(0, 0, 1, .4), rgb(1, 0, 0, .4)), bty = "n")
```

**Interpretation.** Same mean, very different variance. The variance is in *minutes squared*; the standard deviation (the square root) is back in minutes, so we usually quote that.

# Covariance: do two variables move together?

*Experiment:* 200 students. $X$ = hours studied for an exam, $Y$ = exam score. (Simulated: each extra hour adds about 4 points, plus individual noise.)

```{r study-data}
n <- 200
hours <- rnorm(n, mean = 6, sd = 2)
score <- 55 + 4 * (hours - 6) + rnorm(n, 0, 8)
head(data.frame(hours = round(hours, 1), score = round(score, 1)))
```

## The quadrant picture

Draw a vertical line at the average hours and a horizontal line at the average score. The product $(x-\bar x)(y-\bar y)$ is **positive** (blue) in the upper-right and lower-left quadrants and **negative** (red) in the other two.

```{r quadrants}
prod_dev <- (hours - mean(hours)) * (score - mean(score))
plot(hours, score, pch = 16, col = ifelse(prod_dev > 0, "blue", "red"),
     xlab = "Hours studied", ylab = "Exam score",
     main = "Blue: product > 0   Red: product < 0")
abline(v = mean(hours), h = mean(score), lty = 2)
```

Most points are blue, so the *average* product is positive. That average **is** the covariance:

```{r cov-by-hand}
sum(prod_dev) / (n - 1)    # sample covariance by hand
cov(hours, score)          # R's built-in: identical
```

**Interpretation.** A positive covariance means "above-average study time goes with above-average scores". But what does a covariance of about 12 *mean*? "Hour-points" is not a natural unit, so we need correlation.

## Seeing different covariances

Same spread for $X$ and $Y$ in every panel; only the strength and direction of the relationship change.

```{r panels, fig.width=9, fig.height=2.6}
sim_pair <- function(rho, n) {
  x <- rnorm(n); y <- rho * x + sqrt(1 - rho^2) * rnorm(n)
  cbind(x, y)
}
par(mfrow = c(1, 5), mar = c(4, 3, 3, 1))
for (r in c(-0.9, -0.5, 0, 0.5, 0.9)) {
  d <- sim_pair(r, 1000)
  plot(d, pch = 16, cex = .4, col = rgb(0, 0, 0, .3), xlim = c(-4, 4), ylim = c(-4, 4),
       main = sprintf("true rho=%.1f\nsample r=%.2f", r, cor(d[, 1], d[, 2])))
  abline(h = 0, v = 0, lty = 2, col = "red")
}
par(mfrow = c(1, 1))
```

# Correlation: covariance without units

```{r units}
height_in <- rnorm(300, 68, 3)
weight_lb <- 150 + 4 * (height_in - 68) + rnorm(300, 0, 15)
data.frame(
  units = c("inches & pounds", "centimetres & kilograms"),
  covariance  = c(cov(height_in, weight_lb), cov(height_in * 2.54, weight_lb * 0.4536)),
  correlation = c(cor(height_in, weight_lb), cor(height_in * 2.54, weight_lb * 0.4536)))
```

**Interpretation.** Changing units multiplies the covariance by $2.54\times0.4536\approx1.15$, but the correlation does not move. That is why we compare correlations, not covariances, across different variables. A correlation near $\pm1$ means a tight straight-line pattern; near 0 means no *linear* pattern.

# Experiment: two dice

Roll two fair dice. $D_1$, $D_2$ are the dice, $S=D_1+D_2$, and $G=|D_1-D_2|$ is the gap. There are only 36 equally likely outcomes, so we can compute everything **exactly**.

```{r dice}
o <- expand.grid(d1 = 1:6, d2 = 1:6)
o$s <- o$d1 + o$d2; o$gap <- abs(o$d1 - o$d2); o$mx <- pmax(o$d1, o$d2)
res <- rbind(
  "D1 and D2"   = c(pop_cov(o$d1, o$d2),  pop_cor(o$d1, o$d2)),
  "D1 and sum"  = c(pop_cov(o$d1, o$s),   pop_cor(o$d1, o$s)),
  "D1 and max"  = c(pop_cov(o$d1, o$mx),  pop_cor(o$d1, o$mx)),
  "sum and gap" = c(pop_cov(o$s, o$gap),  pop_cor(o$s, o$gap)))
colnames(res) <- c("Cov", "Corr"); round(res, 4)
```

**Interpretation.**

* **D1 and D2:** 0. The dice cannot influence each other.
* **D1 and sum:** $\mathrm{Cov}(D_1,D_1+D_2)=\mathrm{Var}(D_1)+0=35/12\approx2.917$, and $\rho=1/\sqrt2\approx0.707$. Half of the sum's variance comes from the first die.
* **sum and gap:** covariance 0, yet they are **not** independent:

```{r dice-dep}
c(P_joint = mean(o$s == 2 & o$gap == 0),
  P_product = mean(o$s == 2) * mean(o$gap == 0))
```

If the sum is 2, the gap *must* be 0 (joint probability 1/36), but the product of the marginals is only $(1/36)(6/36)$. Uncorrelated $\neq$ independent.

Simulating the experiment gives the same story (with sampling noise):

```{r dice-sim}
a <- sample(6, 10000, TRUE); b <- sample(6, 10000, TRUE)
c(cor_D1_D2 = cor(a, b), cor_D1_sum = cor(a, a + b))
```

# Experiment: guessing on a true/false quiz

A student guesses all 10 questions. $X$ = number correct, $Y$ = number wrong, so $Y=10-X$.

```{r quiz}
correct <- rbinom(5000, 10, 0.5); wrong <- 10 - correct
c(cov = cov(correct, wrong), var_correct = var(correct), cor = cor(correct, wrong))
plot(jitter(correct), jitter(wrong), pch = 16, col = rgb(0, 0, 0, .15),
     xlab = "# correct", ylab = "# wrong", main = "Perfect negative correlation")
cor(correct, 5 * correct)   # points per question: still +1
```

**Interpretation.** $\mathrm{Cov}(X,10-X)=-\mathrm{Var}(X)=-2.5$ and $\rho=-1$ exactly: a perfect decreasing straight line. Multiplying the score by 5 points per question does not change the correlation (+1 with the score).

# Experiment: dealing two cards

$X=1$ if the first card is a heart, $Y=1$ if the second is.

```{r cards}
deck <- rep(c(1, 0), c(13, 39))
draw <- function(replace) deck[sample(52, 2, replace = replace)]
wo <- t(replicate(2e5, draw(FALSE)))
w  <- t(replicate(2e5, draw(TRUE)))
data.frame(sampling = c("with replacement", "without replacement"),
           covariance = c(cov(w[, 1], w[, 2]), cov(wo[, 1], wo[, 2])),
           exact = c(0, -1/272))
```

**Interpretation.** Without replacement, one heart gone leaves fewer hearts in the deck: a **tiny** negative dependence ($\rho=-1/51\approx-0.02$). Dependence can be real yet weak, and it fades as the population grows, which is why polling 1,000 people from millions behaves almost like independent sampling.

# Experiment: commute time and rain (why $\mathrm{Var}(X+Y)$ needs covariance)

Your trip is a bus ride $B$ plus a walk $W$. It rains with probability 0.3 and rain slows **both**:

* $B=20+10R+\varepsilon_B$, $\varepsilon_B\sim N(0,3^2)$
* $W=10+6R+\varepsilon_W$, $\varepsilon_W\sim N(0,2^2)$

```{r commute}
n <- 1e5
rain  <- rbinom(n, 1, 0.3)
bus   <- 20 + 10 * rain + rnorm(n, 0, 3)
walk  <- 10 +  6 * rain + rnorm(n, 0, 2)
total <- bus + walk
c(var_bus = var(bus), var_walk = var(walk), cov = cov(bus, walk), cor = cor(bus, walk))
```

Theory says $\mathrm{Cov}(B,W)=10\cdot6\cdot\mathrm{Var}(R)=60(0.21)=12.6$. Now the variance of the total:

```{r commute-var}
data.frame(
  method = c("Var(B)+Var(W)  (pretend independent)",
             "Var(B)+Var(W)+2Cov(B,W)",
             "Simulated Var(B+W)"),
  variance = c(var(bus) + var(walk),
               var(bus) + var(walk) + 2 * cov(bus, walk),
               var(total)))
```

The second and third lines match; the naive first line is far too small. What does that cost you? Suppose you want to be on time 95% of the time:

```{r commute-buffer}
c(true_95th_percentile  = unname(quantile(total, .95)),
  naive_independent_95th = mean(total) + qnorm(.95) * sqrt(var(bus) + var(walk)))
hist(total, breaks = 60, col = "grey85", main = "Total commute time", xlab = "minutes")
```

**Interpretation.** Because rain hits both parts of the trip, bad days are *bad on both ends* and the total is more spread out (a two-humped distribution: dry days and rainy days). Ignoring covariance gives a buffer that is too small. The same thing happens with money: two stocks that fall together make a riskier portfolio than their separate variances suggest, and two stocks that move oppositely *reduce* the variance of the sum.

To see the effect of the covariance alone, scramble the walk times so they no longer line up with the bus times (same marginal distributions, link broken):

```{r commute-shuffle}
c(original = var(bus + walk), shuffled = var(bus + sample(walk)))
```

# Experiment: temperature and electricity use

A house heats when it's cold and air-conditions when it's hot. $X$ = outdoor temperature minus 20°C (uniform on $[-15,15]$), $Y\approx|X|$ plus noise.

```{r energy}
temp   <- runif(3000, -15, 15)
energy <- abs(temp) + rnorm(3000, 0, 1)
plot(temp, energy, pch = 16, col = rgb(0, 0, 0, .25),
     xlab = "Outdoor temp minus 20C", ylab = "Energy use",
     main = sprintf("Strong relationship, correlation = %.2f", cor(temp, energy)))
c(cov = cov(temp, energy), cor = cor(temp, energy))
```

**Interpretation.** $Y$ is almost determined by $X$, yet the correlation is about 0: cold days (negative $x$) and hot days (positive $x$) both mean high energy use, so the positive and negative products cancel. *Always plot the data.*

# A discrete joint distribution, start to finish

For a student, $X$ = number of late assignments and $Y$ = number of missed lectures, with joint pmf:

```{r pmf}
p <- matrix(c(.2, .1, 0,
              .1, .3, .1,
              0,  .1, .1), nrow = 3, byrow = TRUE,
            dimnames = list(X = 0:2, Y = 0:2))
p
vals <- 0:2
c(sum_check = sum(p), marg_X = rowSums(p), marg_Y = colSums(p))
```

```{r pmf-calc}
EX  <- sum(vals * rowSums(p));  EY  <- sum(vals * colSums(p))
EXY <- sum(outer(vals, vals) * p)
covXY <- EXY - EX * EY
varX  <- sum(vals^2 * rowSums(p)) - EX^2
varY  <- sum(vals^2 * colSums(p)) - EY^2
c(EX = EX, EY = EY, EXY = EXY, Cov = covXY, VarX = varX, VarY = varY,
  Corr = covXY / sqrt(varX * varY))
```

Check by simulating 100,000 students from the table:

```{r pmf-sim}
cells <- expand.grid(Y = vals, X = vals)[, c("X", "Y")]       # X slowest, matches t(p)
idx <- sample(nrow(cells), 1e5, replace = TRUE, prob = as.vector(t(p)))
s <- cells[idx, ]
c(cov = cov(s$X, s$Y), cor = cor(s$X, s$Y))
```

**Interpretation.** $\mathrm{Cov}=1.1-0.9^2=0.29>0$ and $\rho\approx0.59$: students with more late assignments tend to miss more lectures. They are dependent: $P(0,0)=0.2$ but $p_X(0)p_Y(0)=0.09$.

# Correlation is not causation

*Experiment:* daily ice-cream sales and swimming-pool injuries over a summer. Neither causes the other; **temperature** drives both.

```{r confound}
days  <- 120
temp  <- rnorm(days, 28, 5)
icecream <- 50 + 8 * temp + rnorm(days, 0, 25)
injuries <- 1 + 0.25 * temp + rnorm(days, 0, 1.5)
cor(icecream, injuries)                       # clearly positive, and spurious
# Adjust for temperature: correlate what is LEFT after removing the temperature effect
cor(resid(lm(icecream ~ temp)), resid(lm(injuries ~ temp)))
plot(icecream, injuries, pch = 16, col = rgb(0, 0, 0, .4),
     main = "A correlation without causation")
```

**Interpretation.** Once the effect of temperature is removed from both variables, the leftover correlation is near 0, which exposes the lurking variable. (Statisticians call this a partial correlation.)

# Class survey activity (use your own data)

Replace the simulated data below with the class survey (`read.csv("survey.csv")`: `height`, `shoe`, `sleep`, `study`).

```{r survey}
height <- rnorm(60, 67, 4); shoe <- 0.9 * (height - 67) + 9 + rnorm(60, 0, 0.8)
sleep  <- rnorm(60, 7, 1);  study <- pmax(0, 12 - 0.8 * sleep + rnorm(60, 0, 2))
c(height_shoe = cor(height, shoe), sleep_study = cor(sleep, study))
```

*Outliers:* add one badly recorded data point (a student listed as 7 feet tall with size-7 shoes). Pearson correlation is sensitive to it; the rank-based Spearman correlation moves much less.

```{r outlier}
c(pearson_before = cor(height, shoe),
  pearson_after  = cor(c(height, 84), c(shoe, 7)),
  spearman_after = cor(c(height, 84), c(shoe, 7), method = "spearman"))
```

# Summary

```{r summary-table, echo=FALSE}
knitr::kable(data.frame(
  Experiment = c("Study hours / score", "Dice: D1 and sum", "Dice: sum and gap", "True/false quiz",
                 "Two cards, no replacement", "Rain commute", "Temperature / energy"),
  Cov_sign = c("+", "+", "0", "-", "-", "+", "0"),
  Lesson = c("Quadrants: average product of deviations",
             "Part of a variable is shared with the sum",
             "Uncorrelated but dependent",
             "Perfect linear relation: rho = -1; scale-free",
             "Small, real dependence",
             "Var(X+Y) = Var X + Var Y + 2 Cov; risk and buffers",
             "Correlation only sees *linear* association")))
```

# Practice problems

1. If $\mathrm{Var}(X)=4$, $\mathrm{Var}(Y)=9$, $\mathrm{Cov}(X,Y)=3$, find $\mathrm{Var}(X+Y)$, $\mathrm{Var}(X-Y)$ and $\rho$. (Answers: 19, 7, 0.5)
2. In the commute experiment, change the rain probability to 0.1 and 0.5. Which gives the larger covariance between bus and walk? (Use $\mathrm{Var}(R)=p(1-p)$, then check by simulation.)
3. Create two variables that are dependent but have correlation near 0 other than $Y=|X|$.
4. Two independent stocks each have SD \$10. What is the SD of their sum if the correlation is 0? $+0.8$? $-0.8$? (14.1, 19.0, 6.3)
5. Using the dice, compute $\mathrm{Cov}(D_1, \min(D_1,D_2))$ and explain its sign.
